---
output: 
  pdf_document:
    keep_tex: yes
    fig_caption: yes
    number_sections: yes
geometry: margin=2.54cm
title: Experiment Title
subtitle: Web address for GitHub repository
author: Your Name
abstract: Across the country, the expected effects of climate change are being realized in the form of more intense precipitation events and dry periods. In North Texas, the city of Dallas is familiar with both types of extreme hydrologic events, and it has experienced particularly intense events in recent years. An analysis was conducted to determine if floods and droughts, defined as the top and bottom quartiles of daily average discharge, repsectively, have become more frequent in the Trinity River and a tributary near Dallas, TX. The results show that drought discharges have decreased, and flood discharges have slightly increased over time in the Dallas area, and that extreme discharge trends in a larger, more regulated river like the Trinity River are less intense than a lower order tributary.
fontsize: 12pt
mainfont: Times New Roman
editor_options: 
  chunk_output_type: console
---

<Information in these brackets are used for annotating the RMarkdown file. They will not appear in the final version of the PDF document>

\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r include=FALSE}
#knitr::opts_chunk$set(fig.cap = "", fig.path = "Plot")
```

<Note: set up autoreferencing for figures and tables in your document>

```{r setup, include=FALSE, warning=FALSE}
# Set your working directory
setwd("/Users/jakegreif/Environmental_Data_Analytics/EDA_Final_Project")

# Load your packages
library(tidyverse)
library(ggplot2)
library(dplyr)
library(gridExtra)
library(trend)
library(tidyselect)
library(pacman)
library(knitr)
library()

# Set your ggplot theme
mytheme <- theme_bw(base_size = 14) +
  theme(strip.background = element_rect(fill = "white")) +
  theme(plot.title = element_text(hjust=0.5))

theme_set(mytheme)
```


# Research Question and Rationale

Across the country, the expected effects of climate change are being
realized in the form of more intense precipitation events and dry
periods. In North Texas, the city of Dallas is familiar with intense
storms as it resides in a portion of the country where warm air from 
the Gulf of Mexico and cool air flowing down from the Rocky Mountains 
converge, producing dramatic storms. Though, the city is also 
accustomed to dry spells, particularly in the summer. Despite the 
famaliarity this region has with both hydrologic extremes, it has 
experienced particularly intense events in recent years. Therefore, 
this project will examine trends in the flow of the Trinity River 
and a tributary of the Trinity River near Dallas, TX. In particular,
the analysis will attempt to determine if extreme flow rates have 
become more prevalent in recent decades compared to discharges over
the past centry.

Have extreme discharge events become more frequent in the Dallas area?
How do the frequency of extreme discharge events in higher and lower 
order streams in the area compare? To answer this question, data from 
both USGS site 08057000 on the Trinity river in Dallas, TX and USGS 
site 08049500 on the West Fork Trinity River in Grand Prairie, TX will 
be analyzed. Specifically, the mean daily discharge at both sites will 
be characterized based on the frequency of occurance. As per USGS 
percentile thresholds, a discharge that falls above the 75th percentile
is considered to be above normal, and a discharge that falls below the 
25th percentile is considered below normal. Therefore, in the context 
of this analysis, a discharge above the 75th percentile is considered 
a "flood", and a discharge below the 25th percentile is considered a 
"drought." The goal of this analysis is to determine if recent extreme 
events are indicative of an emerging trend, as well as if a larger,
more regulated river (i.e.the Trinity River) has less frequent extreme
discharge than a lower order tributary.
\newpage

# Dataset Information

The dataset used in the analysis was sourced from the U.S. Geological
Survey's (USGS) Water Data website. While there is a collection of
physical and chemical data available, the only data used for this
analysis is average daily discharge (ft^3/s). The gauge sites used in
the analysis were found using an interactive map on the USGS Water Data
website, and were selected based on their proximity to the city of
Dallas. The West Fork Trinity River was chosen as the representative
tributary due the lack of development surrounding the river, as well as
the lack of river control present upstream. After idenifying the gauge
station number at both locations, the number was entered into the
'Site Number' search bar. Then, the 'Discharge' check-box was selected
and the years 1925-2019 were selected. 1925 was chosen as the start
date based on the earliest available data for both datasets. Both
datasets also contain maximum and minimum daily discharge, however that
data is only available beginning in 1989 for the Trinity River, and in
2001 for the West Fork Trinity River. Therefore, using the average
daily discharge allowed the analysis to encompass a larger timeframe. 

agency_id   |  site_no     |  datetime    |  discharge.mean
----------- |  ----------  |  ----------- |  ----------------
USGS        |  8057000     |  1925-04-01  |  31.0
USGS        |  8057000     |  1925-04-02  |  30.0
USGS        |  8057000     |  1927-04-03  |  32.0

\newpage

# Exploratory Data Analysis and Wrangling

```{r warning=FALSE, results='markup'}
## Data Wrangling
trinity.raw <- 
  read.csv("../EDA_Final_Project/Data/Raw_Data/USGS_Site08057000_Flow_Raw.csv")
westfork.raw <- 
  read.csv("../EDA_Final_Project/Data/Raw_Data/USGS_Site08049500_Flow_Raw.csv")

class(trinity.raw)
class(westfork.raw)

colnames(trinity.raw)
colnames(westfork.raw)

colnames(trinity.raw) <- c("agency_cd", "site_no", "datetime", 
                              "discharge.max", "discharge.max.approval", 
                              "discharge.min", "discharge.min.approval", 
                              "discharge.mean", "discharge.mean.approval")
colnames(westfork.raw) <- c("agency_cd", "site_no", "datetime", 
                              "discharge.max", "discharge.max.approval", 
                              "discharge.min", "discharge.min.approval", 
                              "discharge.mean", "discharge.mean.approval")

head(trinity.raw)
head(westfork.raw)

class(trinity.raw$datetime)
class(westfork.raw$datetime)

# Convert "datetime" class to date
trinity.raw$datetime <- as.Date(trinity.raw$datetime, 
                                               format = "%m/%d/%y")
westfork.raw$datetime <- as.Date(westfork.raw$datetime, 
                                               format = "%m/%d/%y")

# Fix years before 1925
trinity.raw$datetime <- format(trinity.raw$datetime, "%y%m%d")
westfork.raw$datetime <- format(westfork.raw$datetime, "%y%m%d")

create.early.dates <- (function(d) {
       paste0(ifelse(d > 190404,"19","20"),d)
       })

trinity.raw$datetime <- create.early.dates(trinity.raw$datetime)
westfork.raw$datetime <- create.early.dates(westfork.raw$datetime)

# Reformat dates
trinity.raw$datetime <- as.Date(trinity.raw$datetime, 
                                               format = "%Y%m%d")
westfork.raw$datetime <- as.Date(westfork.raw$datetime, 
                                               format = "%Y%m%d")

# Convert "site_no" class to character
trinity.raw$site_no <- as.character(trinity.raw$site_no)
westfork.raw$site_no <- as.character(westfork.raw$site_no)

# Separate "datetime" column
trinity.date <- 
  separate(trinity.raw, datetime, c("Year", "Month", "Day"))
westfork.date <- 
  separate(westfork.raw, datetime, c("Year", "Month", "Day"))

# Thin data to contain only mean discharge
trinity.thinned <- select(trinity.date, site_no, Year, discharge.mean)
westfork.thinned <- select(westfork.date, site_no, Year, discharge.mean)

# Remove "discharge.mean" NAs
trinity.flow <- na.omit(trinity.thinned)
westfork.flow <- na.omit(westfork.thinned)

# Merge datasets
all.flow.data <- merge(trinity.flow, westfork.flow, all = TRUE)
```

The above R chunk contains code wrangles and tidies the datasets. After the datasets are imported, the dates are not formatted correctly and need to be converted to the proper class, as well as reformated to be usable for the analysis. The datsets are then refined to be in a useful format for the analysis.

```{r warning=FALSE}
# Determine extreme event thresholds
quantile(trinity.flow$discharge.mean)
quantile(westfork.flow$discharge.mean)

# Create flood/drought data frames
trinity.flood <- filter(trinity.flow, discharge.mean >= 1580)
trinity.drought <- filter(trinity.flow, discharge.mean <= 250)
westfork.flood <- filter(westfork.flow, discharge.mean >= 512.25)
westfork.drought <- filter(westfork.flow, discharge.mean <= 115)

# Create count column of frequency of extreme events by year
trinity.flood.freq <- count(trinity.flood, Year)
trinity.drought.freq <- count(trinity.drought, Year)
westfork.flood.freq <- count(westfork.flood, Year)
westfork.drought.freq <- count(westfork.drought, Year)

# Append "site_no" column to frequency dataframes
trinity.flood.freq$site_no <- c(08057000)
trinity.drought.freq$site_no <- c(08057000)
westfork.flood.freq$site_no <- c(08049500)
westfork.drought.freq$site_no <- c(08049500)

# Rearrange columns
trinity.flood.freq <- 
  select(trinity.flood.freq, site_no, Year, n)
trinity.drought.freq <- 
  select(trinity.drought.freq, site_no, Year, n)
westfork.flood.freq <- 
  select(westfork.flood.freq, site_no, Year, n)
westfork.drought.freq <- 
  select(westfork.drought.freq, site_no, Year, n)

# Make "Year" class numeric
trinity.flood.freq$Year <- as.numeric(trinity.flood.freq$Year)
trinity.drought.freq$Year <- as.numeric(trinity.drought.freq$Year)
westfork.flood.freq$Year <- as.numeric(westfork.flood.freq$Year)
westfork.drought.freq$Year <- as.numeric(westfork.drought.freq$Year)

# Create all flood and all drought dataframes
all.flood.data <- merge(trinity.flood.freq, westfork.flood.freq, all = TRUE)
all.drought.data <- merge(trinity.drought.freq, westfork.drought.freq, all = TRUE)
  # Delete this stuff??
```

New dataframes of flood/drought occurance by year are created for the analysis. The top and bottom quartiles are determined for each gauge, and the dataframe is further refined in order to create the most simple dataset possible for the analysis.

```{r Exploratory, warning=FALSE, echo=FALSE}
## Exploratory Graphsj
Fig1 <- ggplot(NULL, aes(x = discharge.mean)) +
  geom_histogram(data = trinity.flow, fill = "red") +
  geom_histogram(data = westfork.flow, fill = "blue", alpha = 0.5) +
  ylim(0,1500) +
  xlim(0,20000)
 
Fig2 <- ggplot(all.flow.data) +
  geom_boxplot(aes(x = site_no, y = discharge.mean)) +
  ylim(0,2000)

print(Fig1)
print(Fig2)

Fig3 <- ggplot(trinity.flood.freq, aes(x = Year, y = n)) +
  geom_point() +
  geom_smooth(method = lm) +
  ylab("Frequency") +
  ggtitle("Trinity River Flood Occurance") +
  theme(plot.title = element_text(size = 12))

Fig4 <- ggplot(trinity.drought.freq, aes(x = Year, y = n)) +
  geom_point() +
  geom_smooth(method = lm, color = "red") +
  ylab("Frequency") +
  ggtitle("Trinity River Drought Occurance") +
  theme(plot.title = element_text(size = 12))

Fig5 <- ggplot(westfork.flood.freq, aes(x = Year, y = n)) +
  geom_point() +
  geom_smooth(method = lm) +
  scale_y_continuous(limits = c(0,250)) +
  ylab("Frequency") +
  ggtitle("West Fork Trinity River Flood Occurance") +
  theme(plot.title = element_text(size = 12))

Fig6 <- ggplot(westfork.drought.freq, aes(x = Year, y = n)) +
  geom_point() +
  geom_smooth(method = lm, color = "red") +
  scale_y_continuous(limits = c(0,300)) +
  ylab("Frequency") +
  ggtitle("West Fork Trinity River Drought Occurance") +
  theme(plot.title = element_text(size = 12)) 

grid.arrange(Fig3, Fig4, Fig5, Fig6)

```

The exploratory graphs help orient the data, and it confirms assumptions about the data, such as the tributary river having a similar trend in discharge over time, but at a less intense magnitude. It also reveals some unexpected trends that will be tested in the statistical analysis. In particular, it reveals the trend of decreasing "drought" days and slightly increasing "flood" days.

\newpage

# Analysis
```{r warning=FALSE, error=FALSE}
# Run Man-Kendall test to test for monotonic relationship
mk.test(trinity.flood.freq$n)
  # p < 0.05, trend present (z = +2.21)
mk.test(trinity.drought.freq$n)
  # p < 0.05, trend present (z = -2.35)
mk.test(westfork.flood.freq$n)
  # p < 0.05, trend present (z = +2.13)
mk.test(westfork.drought.freq$n)
  # p < 0.05, trend present (z = -5.75)

# Run Pettitt's test 
pettitt.test(trinity.flood.freq$n)
  # Change point @ row 56 (1980)
pettitt.test(trinity.drought.freq$n)
  # Change point @ row 40 (1964)
pettitt.test(westfork.flood.freq$n)
  # Change point @ row 64 (1988)
pettitt.test(westfork.drought.freq$n)
  # Change point @ row 39 (1964)

# Run separate Man-Kendall for each change point
mk.test(trinity.flood.freq$n[1:55])
  # p > 0.05, no trend
mk.test(trinity.flood.freq$n[56:95])
  # p > 0.05, no trend
mk.test(trinity.drought.freq$n[1:39])
  # p > 0.05, no trend
mk.test(trinity.drought.freq$n[40:61])
  # p < 0.05, trend present (z = -2.09)
mk.test(westfork.flood.freq$n[1:63])
  # p > 0.05, no trend
mk.test(westfork.flood.freq$n[64:95])
  # p > 0.05, no trend
mk.test(westfork.drought.freq$n[1:38])
  # p > 0.05, no trend
mk.test(westfork.drought.freq$n[39:66])
  # P < 0.05, trend present (z = -3.42)

# Test for second change point
pettitt.test(trinity.drought.freq$n[40:61])
  # p > 0.05, no change point
pettitt.test(westfork.drought.freq$n[39:66])
  # p < 0.05, change point @ row 44 (1969)

# Second MK test for second change point
mk.test(westfork.drought.freq$n[39:43])
mk.test(westfork.drought.freq$n[44:66])
  # p > 0.05 for both, no trends

# Run Spearman's Rho test (non-parametric)
cor.test(trinity.flood.freq$n, trinity.flood.freq$Year, 
         method = "spearman", exact = FALSE)
 # p < 0.05 (0.02), rho = 0.238, weak correlation
cor.test(trinity.drought.freq$n, trinity.drought.freq$Year, 
         method = "spearman", exact = FALSE)
 # p < 0.05 (0.02), rho = -0.295, weak correlation
cor.test(westfork.flood.freq$n, westfork.flood.freq$Year, 
         method = "spearman", exact = FALSE)
  # p < 0.05 (0.04), rho = 0.214, weak correlation
cor.test(westfork.drought.freq$n, westfork.drought.freq$Year, 
         method = "spearman", exact = FALSE)
  # p < 0.05 (3.2^-10), rho = -0.681, strong correlation
```

First, a Mann_Kendall test was run to determine if the trends in discharge are monotonic. A Pettitt's test was then run on each dataset to see if there were any change points present, which could be indicative of increased river control and/or dramatic climatic changes. Because all four trends are monotonic, albeit weakly monotonic, a Spearman Rho's test was run to determine the strength of the correlation between the extreme hydrologic events and time. 

```{r Visual, warning=FALSE, echo=FALSE}
# Visualization Graphs

# 1. Scatter plots of data with upper/lower 25% thresholds filled.
Fig7 <- ggplot(trinity.raw, aes(x = datetime, y = discharge.mean)) +
  geom_rect(xmin = as.Date("1925-01-01"), xmax = as.Date("2020-01-01"), 
      ymin = 0, ymax = 250, fill = "darkgoldenrod", alpha = 0.75) +
  geom_rect(xmin = as.Date("1925-01-01"), xmax = as.Date("2020-01-01"), 
      ymin = 1580, ymax = 2000, fill = "dodgerblue3", alpha = 0.75) +
  geom_point(color = "gray41", alpha = 0.2) +
  scale_y_continuous(limits = c(0,2000)) +
  geom_text(x = as.Date("2019-01-01"), y = 50,
      label = "Drought", hjust = 1, fontface = "bold") +
  geom_text(x = as.Date("2019-01-01"), y = 1950,
      label = "Flood", hjust = 1, fontface = "bold") +
  ylab(expression("Mean Discharge" ~ (ft^{3}/~"s"))) +
  xlab("Year") +
  ggtitle("Trinity River Mean Daily Discharge") +
  theme(plot.title = element_text(size = 14))

print(Fig7)

Fig8 <- ggplot(westfork.raw, aes(x = datetime, y = discharge.mean)) +
  geom_rect(xmin = as.Date("1925-01-01"), xmax = as.Date("2020-01-01"), 
      ymin = 0, ymax = 115, fill = "darkgoldenrod", alpha = 0.75) +
  geom_rect(xmin = as.Date("1925-01-01"), xmax = as.Date("2020-01-01"), 
      ymin = 512.25, ymax = 750, fill = "dodgerblue3", alpha = 0.75) +
  geom_point(color = "gray41", alpha = 0.2) +
  scale_y_continuous(limits = c(0,750)) +
  geom_text(x = as.Date("2019-01-01"), y = 50,
      label = "Drought", hjust = 1, fontface = "bold") +
  geom_text(x = as.Date("2019-01-01"), y = 725, 
      label = "Flood", hjust = 1, fontface = "bold") +
  ylab(expression("Mean Discharge" ~ (ft^{3}/~"s"))) +
  xlab("Year") +
  ggtitle("West Fork Trinity River Mean Daily Discharge") +
  theme(plot.title = element_text(size = 12))

print(Fig8)
```
Figures 1 and 2 visually dipict how  often the discharges of both rivers fall into the top and bottom quintiles, characterized as "Drought" for the bottom 25% and "Flood" for the top 25%. Both of the above figures were rescaled to better visualize the thresholds for flood and drought discharges for each river. 7,615 values above 2,000 were removed from Figure 2, and 6,645 values above 750 were removed from Figure 3.

```{r Visual2, warning=FALSE, echo=FALSE}
# Visualization graphs

# 2. Trendline w/ break point floods 
Fig9 <- ggplot(trinity.flood.freq, aes(x = Year, y = n)) +
  geom_point(position = position_jitter(height = 0L, seed = 1L),
      size = 0.5, color = "gray41", alpha = 0.25) +
  geom_linerange(aes(x=Year, ymin = 0, ymax = n),
      position = position_jitter(height = 0L, seed = 1L),
      size = 1, color = "gray41", alpha = 0.5) +
  geom_vline(xintercept = as.numeric("1980"),
      color = "red3", lty = 5, size = 1, alpha = 0.75) +
  geom_smooth(method = lm, se = F,
      color = "dodgerblue3", alpha = 0.75) +
  ylab("") +
  xlab("") +
  scale_y_continuous(limits = c(0,250), expand = c(0,0)) +
  ggtitle("Trinity River Flood Occurance") +
  theme(plot.title = element_text(size = 12))

Fig10 <- ggplot(westfork.flood.freq, aes(x = Year, y = n)) +
  geom_point(position = position_jitter(height = 0L, seed = 1L),
      size = 0.5, color = "gray41", alpha = 0.25) +
  geom_linerange(aes(x=Year, ymin = 0, ymax = n),
      position = position_jitter(height = 0L, seed = 1L),
      size = 1, color = "gray41", alpha = 0.5) +
  geom_vline(xintercept = as.numeric("1988"),
      color = "red3", lty = 5, size = 1, alpha = 0.75) +
  geom_smooth(method = lm, se = F,
      color = "dodgerblue3", alpha = 0.75) +
  scale_y_continuous(limits = c(0,250), expand = c(0,0)) +
  ylab("Frequency") +
  ggtitle("West Fork Trinity River Flood Occurance") +
  theme(plot.title = element_text(size = 12))

grid.arrange(Fig9, Fig10)

# 3. Trendline w/ break point droughts
Fig11 <- ggplot(trinity.drought.freq, aes(x = Year, y = n)) +
  geom_point(position = position_jitter(height = 0L, seed = 1L),
      size = 0.5, color = "gray41", alpha = 0.25) +
  geom_linerange(aes(x=Year, ymin = 0, ymax = n),
      position = position_jitter(height = 0L, seed = 1L),
      size = 1, color = "gray41", alpha = 0.5) +
  geom_vline(xintercept = as.numeric("1964"),
      color = "red3", lty = 5, size = 1, alpha = 0.75) +
  geom_smooth(method = lm, se = F,
      color = "darkgoldenrod", alpha = 0.75) +
  ylab("") +
  xlab("") +
  scale_y_continuous(limits = c(0,250), expand = c(0,0)) +
  ggtitle("Trinity River Drought Occurance") +
  theme(plot.title = element_text(size = 12))

Fig12 <- ggplot(westfork.drought.freq, aes(x = Year, y = n)) +
  geom_point(position = position_jitter(height = 0L, seed = 1L),
      size = 0.5, color = "gray41", alpha = 0.25) +
  geom_linerange(aes(x=Year, ymin = 0, ymax = n),
      position = position_jitter(height = 0L, seed = 1L),
      size = 1, color = "gray41", alpha = 0.5) +
  geom_vline(xintercept = as.numeric("1964"),
      color = "red3", lty = 5, size = 1, alpha = 0.75) +
  geom_smooth(method = lm, se = F,
      color = "darkgoldenrod", alpha = 0.75) +
  scale_y_continuous(limits = c(0,250), expand = c(0,0)) +
  scale_x_continuous(limits = c(1925,2019)) +
  ylab("Frequency") +
  ggtitle("West Fork Trinity River Drought Occurance") +
  theme(plot.title = element_text(size = 12))

grid.arrange(Fig11, Fig12)
```

Figures 3 and 4 drive home the main takeaway from the analysis- "drought" discharges have decreased, and "flood" discharges have slightly increased over time in the Dallas area. Red dashed lines indicate change points, which provide additional information about when infrastructural or climatic changes began to really affect the surface water hydrology of the Dallas area.

\newpage

# Summary and Conclusions

The discharge of rivers in the Dallas area show that the region is
getting wetter over time. In both the Trinity River and the West Fork
Trinity River, "droughts" have become less frequent, while "floods"
have become more frequent. However, even though the Mann-Kendall tests
revealed that these trends are indeed monotonic, none of the trends are
particularlly strong. The strongest correlation between extreme
discharge frequency and time were "drought" discharge conditions in the
West Fork Trinity River, which can be characterized as a moderately
strong negative trend based on the results of the Man-Kendall test
(z = -5.75). The other Man-Kendall tests produced absolute z values no
greater than 2.35, indiicating that while the trends are monotonic,
they are relatively weak trends. 

Pettitt's test was used to determine if any change points exist within
the data. A change point is where the magnitude or direction of the
trend changes in the data. The analysis found that in both rivers, a
change point in "flood" discharges occurred in the 1980s, and the
change point in the "drought" discharges occurred in 1964. The only
significant trends before or after the change point were found in
the "drought" datasets for both rivers, which experienced negative
trends after 1964. Meaning, both the Trinity River and the West Fork
Trinity River experienced less frequent "drought" discharges after
1964. 

The results of this analysis indicate that the Trinity River and it's
tributary, the West Fork Trinity River, are experiencing similar trends
in the frequency of extreme hydrologic events. However, the trends in
the larger Trinity River are weaker overall than the West Fork Trinity
River, particularlly in "drought" discharge frequency. Maintaing
minimum flows in the Trinity River is much more important than
controling flows in the West Fork Trinity River, considering the
Trinity River is the largest river in the area and is used as a source
of drinking water. The analysis confirms that recent extreme events in
the Dallas area are indicative of an emerging trend. It also confirms
that extreme discharge trends in a larger, more regulated river like
the Trinity River are less intense than a lower order tributary.
